Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-11-24 tex/manuscript.tex
Lagt inn Eivinds korreksjoner i manus

The two factors both cause a bias in the $F$-statistic. The first factor, $\tilde q\tilde\sigma^2/(M-1)\sigma^2$, captures the unbalancedness of the design, but is otherwise independent of the sample size. The second factor, $1-(m-1)/(n-M-m+1)$, is due to loss of variance from centering of the data, which can be influential when there are few samples in each batch. Finally, the degrees of freedom, $\tilde q$ and $n-M-m+1$, of the F-distribution are smaller than assumed by the $\text{F}_{M-1,n-M}$ distribution, and so the variance is larger.

@@

Knowing that adjustment for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust published results where such a method has been applied? Essentially, when the batch--group configuration is balanced, batch adjustment will not bias the group difference estimates. For unbalanced designs, however, batch effects may act as a confounder, and a re-analysis without relying on the two-step approach is the most rigorous path. However, such a re-analysis may be infeasible or impractical, depending on the availability of the raw data and proper method description, the ability of analysis tools to handle batch effects. To reach a reliable result, batch effects need to be handled in some way or another, and the necessary statistics and bioinformatics resources.

>>>>

The two factors both cause a bias in the $F$-statistic. The first factor, $\tilde q\tilde\sigma^2/(M-1)\sigma^2$, captures the degree of unbalance of the design, but is otherwise independent of the sample size. The second factor, $1-(m-1)/(n-M-m+1)$, is due to loss of variance from centering of the data, which can be influential when there are few samples in each batch. Finally, the degrees of freedom, $\tilde q$ and $n-M-m+1$, of the F-distribution are smaller than assumed by the $\text{F}_{M-1,n-M}$ distribution, and so the variance is larger.

@@

Knowing that adjustment for batch effects while preserving the group difference may lead to varying degree of false results, to what extent can an investigator trust published results where such a method has been applied? Essentially, when the batch--group configuration is balanced, batch adjustment will not bias the group difference estimates. For unbalanced designs, however, batch effects may act as a confounder, and a re-analysis without relying on the two-step approach is the most rigorous path. However, such a re-analysis may be infeasible or impractical, depending on the availability of the raw data and proper method description, and on the ability of analysis tools to handle batch effects. To reach a reliable result, batch effects need to be handled in some way or another, and the necessary statistics and bioinformatics resources.
